---
title: "Transcription and Splicing Dynamics"
author: "Edward Wallace, ewjwallace@gmail.com"
website: https://ewallace.github.io
date: "11th January 2018"
output:
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup,warning=FALSE,message=FALSE,echo=FALSE}
## knitr options for report generation
knitr::opts_chunk$set(warning=FALSE,message=FALSE,echo=FALSE,cache=FALSE,
                      results="show")
library(tidyverse)

theme_set(theme_bw(base_size=12) + 
              theme(strip.background=element_blank()))
```

This document uses a simple first-order model of transcription, splicing, and decay, to ask what changes in transcription and splicing rate we could reasonably detect. It concludes that the fraction unspliced should be a good way to detect changes in splicing rate.

I wrote this in 2017 to clarify some discussions in the Beggs lab, and shared it on github in 2018. Please let me know if you find it interesting or useful.

## The model

Pre-mRNA (P) is **transcribed** at rate $\tau$, **spliced** at rate $\sigma$ to mRNA (M), which is then **degraded** at rate $\lambda$. As usual in these models, $\phi$ is used as a placeholder for emptiness - creation or decay.

\[ \phi \xrightarrow{\tau} P \xrightarrow{\sigma} M \xrightarrow{\lambda} \phi
\]

In differential equations,

\[ \frac{dP}{dt} = \tau - \sigma P \]

\[ \frac{dM}{dt} = \sigma P - \lambda M \]

This model **ignores** degradation of pre-mRNA.

### Simulate from model

Here we simulate using the exact analytical solution of the simple model.

We use $\tau=1$, $\sigma=1$, $\lambda=0.1$. I.E. rate of splicing per pre-mRNA molecule is 10X rate of decay per mRNA.

We start with initial conditions $P=0$, $M=0$, which approximates a perfect pulse-labeling experiment, for example with 4tU.

```{r model1}
sim1 <- function(tau=1, sigma=1, lambda=0.1, P0=0, M0=0,
                 timevec=seq(0,20,length.out = 2001)) {
    Ps = tau / sigma
    Ms = tau / lambda
    Mfrac = sigma * ( P0 - Ps ) / (lambda - sigma)
    tibble( timeval = timevec,
            Pt      = ( P0 - Ps ) * exp( - sigma * timevec ) + Ps,
            Mt      = Mfrac * exp( - sigma * timevec ) + 
                ( M0 - Ms - Mfrac ) * exp ( - lambda * timevec ) + 
                Ms)
}
v1 <- sim1()

ggplot(data=v1,aes(x=timeval)) +
    geom_line(aes(y=Pt,colour="pre-mRNA")) +
    geom_line(aes(y=Mt,colour="mRNA")) +
    scale_x_continuous("time (min)",expand=c(0.001,0.001)) + 
    scale_y_continuous("RNA quantity",limits=c(0,10),expand=c(0.001,0.001)) 
```

### Vary transcription and splicing rate

Here with varying $\tau$ and $\sigma$, keeping fixed $\lambda=0.1$.

This shows that the amount of pre-mRNA rapidly converges to the expected steady-state value of $\tau/\sigma$, and the amount of mRNA more slowly converges to $\tau/\lambda$.

```{r simgrid}
simgrid <- expand.grid(tau=c(0.5,1,2),sigma=c(0.5,1,2)) %>%
    group_by(tau,sigma) %>%
    do(sim1(.$tau[1],.$sigma[1],lambda = 0.1)) %>%
    mutate(taufac=paste("τ =",tau),
           sigmafac=paste("σ =",sigma))

ggplot(data=simgrid,aes(x=timeval)) +
    geom_line(aes(y=Pt,colour="pre-mRNA")) +
    geom_line(aes(y=Mt,colour="mRNA")) +
    scale_x_continuous("time (min)",expand=c(0.01,0.01)) + 
    scale_y_continuous("RNA quantity",expand=c(0.01,0.01)) +
    facet_grid(sigmafac~taufac) 

```

### Zooming in on initial 5 minutes 

We zoom in on the initial times to see changes from varying transcription and splicing rate. Same curves as above, varying $\tau$ and $\sigma$, keeping fixed $\lambda=0.1$.

```{r simgrid_zoom}
ggplot(data=simgrid,aes(x=timeval)) +
    geom_line(aes(y=Pt,colour="pre-mRNA")) +
    geom_line(aes(y=Mt,colour="mRNA")) +
    scale_x_continuous("time (min)",expand=c(0.01,0.01),limits=c(0,5)) + 
    scale_y_continuous("RNA quantity",expand=c(0.01,0.01),limits=c(0,5)) +
    facet_grid(sigmafac~taufac) 
```

### The pre-mRNA level alone barely distinguishes splicing rate from transcription rate

```{r simgrid_premRNA,fig.width=5,fig.height=3.5}
ggplot(data=simgrid %>% dplyr::filter(tau==1),
       aes(x=timeval,colour=sigma,group=sigma)) +
    geom_line(aes(y=Pt)) +
    scale_x_continuous("time (min)",expand=c(0.001,0.001),limits=c(0,5)) + 
    scale_y_continuous("pre-mRNA level, tau = 1",expand=c(0.001,0.001),limits=c(0,1)) +
    scale_color_continuous("splice rate σ",low="red",high="black") 

ggplot(data=simgrid %>% dplyr::filter(sigma==1),
       aes(x=timeval,colour=tau,group=tau)) +
    geom_line(aes(y=Pt)) +
    scale_x_continuous("time (min)",expand=c(0.001,0.001),limits=c(0,5)) + 
    scale_y_continuous("pre-mRNA level, sigma =1",expand=c(0.001,0.001),limits=c(0,1)) +
    scale_color_continuous("transcription\n rate τ",low="cyan",high="black") 
```

Because changing the splicing rate changes the rate of approach to steady-state, in principle we could detect this with a fast enough time-course.

### Fraction unspliced detects the splicing rate

Again with varying $\tau$ and $\sigma$, keeping fixed $\lambda=0.1$.

In this model, the unspliced/total ratio is **independent** of transcription rate $\tau$.

```{r simgrid_fracspliced,fig.width=5,fig.height=3.5}
ggplot(data=simgrid %>% dplyr::filter(tau==1),
       aes(x=timeval,colour=sigma,group=sigma)) +
    geom_line(aes(y=Pt/(Pt + Mt))) +
    scale_x_continuous("time (min)",expand=c(0.001,0.001),limits=c(0,5)) + 
    scale_y_continuous("Fraction unspliced: pre / ( pre + mRNA)",expand=c(0.001,0.001),limits=c(0,1)) +
    scale_color_continuous("splice rate σ",low="red",high="black") 
```

### Initial fraction unspliced is not very sensitive to changes in mRNA decay rate

Here varying $\sigma$ and $\lambda$, keeping fixed transcription rate $\tau=1$.

```{r decay_fracspliced,fig.width=5,fig.height=3.5}
simgridl <- expand.grid(tau=1,sigma=c(0.5,1,2),lambda=c(0.05,0.1,0.2)) %>%
    group_by(sigma,lambda) %>%
    do(sim1(.$tau[1],.$sigma[1],.$lambda[1])) %>%
    mutate(#taufac=paste("τ =",tau),
           sigmafac=paste("σ =",sigma),
           lambdafac=paste("λ =",lambda))

ggplot(data=simgridl,
       aes(x=timeval,colour=sigma,linetype=lambdafac,
           group=interaction(sigma,lambda))) +
    geom_line(aes(y=Pt/(Pt + Mt))) +
    scale_x_continuous("time (min)",expand=c(0.001,0.001),limits=c(0,5)) + 
    scale_y_continuous("Fraction unspliced: pre / ( pre + mRNA)",expand=c(0.001,0.001),limits=c(0,1)) +
    scale_color_continuous("splice rate σ",low="red",high="black") + 
    scale_linetype_discrete("decay rate λ") 

```

This point applies to the initial part of the timecourse, which we could detect by pulse-labeling. The same  approach shows that, if mRNA splicing rate were constant and mRNA decay rate variable, then **steady-state** fraction unspliced would be a good proxy for decay rate. I heard this point made by Uwe Ohler at a conference in 2015.


## Conclusion

I conclude that the fraction unspliced should be a good measure of splicing rate as opposed to transcription, as long as

  - the first-order model is close enough to reality
  - background is low (we did not check this directly, yet)
  - mature mRNA decay rate is much less than splicing rate and does not vary much (2x is ok)

Further, this suggests that fraction unspliced is a good proxy **even at steady state**. Pulse-labeling will help to resolve the dynamics, including effects of promoter on transcription and decay rate. 



#### Acknowledgments

Thanks to Guido Sanguinetti, Jean Beggs, and David Barrass for discussions.